Hartree-Fock Stability of the Electron Gas

### 2017-10-04 doc/presentations/Fall_2017/Fall2017.tex
Added presentation for Norskov lunch

\title{Hartree-Fock Stability}
\subtitle{Applications to the Homogeneous Electron Gas}
\date{\today}
\author{Evan Curtin}
\institute{University of Illinois at Urbana-Champaign}
\titlegraphic{\hfill\includegraphics[height=1.5cm]{../../images/imark_bold.eps}}

@@

\begin{frame}{Matrix Factorizations}
\vspace{-6mm}
\begin{eqnarray*}
  \begin{bmatrix}
    \mathbf{A} & \mathbf{B} \\
    \mathbf{B^*} & \mathbf{A^*} 
  \end{bmatrix}
  \begin{bmatrix}
    \mathbf{d} \\
    \mathbf{d^*} 
  \end{bmatrix}
  = 2E_2
  \begin{bmatrix}
    \mathbf{d} \\
    \mathbf{d^*} 
  \end{bmatrix}
\end{eqnarray*}
We can now apply the similarity transform defined by the Unitary matrix 
\begin{eqnarray*}
  \mathbf{U} = \frac{1}{\sqrt{2}}
  \begin{bmatrix}
    \mathbf{I} & -\mathbf{I} \\
    \mathbf{I} & \mathbf{I} 
  \end{bmatrix}
\end{eqnarray*}
after which the transformed eigenvalue problem has the form
\begin{eqnarray*}
  \frac{1}{2}
  \begin{bmatrix}
    \mathbf{A + B + A^* + B^*} & \mathbf{-A + A^* + B - B^*} \\
    \mathbf{-A + A^* - B + B^*} & \mathbf{A^* + A - B - B^*} 
  \end{bmatrix}
  \begin{bmatrix}
    \mathbf{d + d^*} \\
    \mathbf{d - d^*} 
  \end{bmatrix}
  &=& 2E_2
  \begin{bmatrix}
    \mathbf{d + d^*} \\
    \mathbf{-d + d^*} 
  \end{bmatrix} \\
  &=& 2E_2
  \begin{bmatrix}
    \mathbf{Re(d)} \\
    \mathbf{Im(d)} 
  \end{bmatrix}
\end{eqnarray*}
If $\mathbf{A}$ and $\mathbf{B}$ are both real, $\mathbf{A = A^*}$ and $\mathbf{B = B^*}$ and the 
above simplifies to 
\begin{eqnarray*}
  \begin{bmatrix}
    \mathbf{A + B} & \mathbf{0} \\
    \mathbf{0} & \mathbf{A - B} 
  \end{bmatrix}
  \begin{bmatrix}
    \mathbf{Re(d)} \\
    \mathbf{Im(d)} 
  \end{bmatrix}
  &=& 2E_2   
  \begin{bmatrix}
    \mathbf{Re(d)} \\
    \mathbf{Im(d)} 
  \end{bmatrix}
\end{eqnarray*}
\end{frame}

{%
	\setbeamertemplate{frame footer}{1) Seeger, R.; Pople, J. A. J. Chem. Phys. 1977, 66 (7), 3045.}
\begin{frame}{Hartree-Fock Stability Conditions}
	\begin{itemize}
		\item[]{Doing this for all cases yields} \\~\\
		\item[]{
			\resizebox{0.9\textwidth}{!}{%
			\begin{tabular}{r@{\qquad}cccccc}
				  \multirow{1}{*}{Solution Type} & \multicolumn{6}{c}{Space Type}  \\
				  \cmidrule{2-7}
				  & Real RHF & Complex RHF & Real UHF & Complex UHF & Real GHF & Complex GHF \\
				  \midrule
				  Real RHF    & ${}^1\mathbf{A}' + {}^1\mathbf{B}'$ & ${}^1\mathbf{A}' - {}^1\mathbf{B}'$ & 
				  ${}^3\mathbf{A}' + {}^3\mathbf{B}'$ & ${}^3\mathbf{A}' - {}^3\mathbf{B}'$ & 
				  ${}^3\mathbf{A}' + {}^3\mathbf{B}'$ & ${}^3\mathbf{A}' - {}^3\mathbf{B}'$  \\
				  Complex RHF & - & \alert{${}^1\mathbf{H}'$} & - & \alert{${}^3\mathbf{H}'$} & - & 
				  ${}^3\mathbf{H}'$ \\
				  Real UHF    & - & - & $\mathbf{A}' + \mathbf{B}'$ & $\mathbf{A}' - \mathbf{B}'$ & 
				  $\mathbf{A}'' + \mathbf{B}''$ & $\mathbf{A}'' - \mathbf{B}''$ \\
				  Complex UHF & - & - & - &  $\mathbf{H}'$ & - & $\mathbf{H}'$ \\
				  Real GHF    & - & - & - & - & $\mathbf{A} - \mathbf{B}$ &  $\mathbf{A} - \mathbf{B}$  \\
				  Complex GHF & - & - & - & - & - & $\mathbf{H}$ \\
				  \bottomrule
			\end{tabular}}
		\item[] \tiny Table reproduced from Seeger \& Pople$^1$
		}
	\end{itemize}
\end{frame}

@@

@@

 2. Li, R.-C.; Zhang, L.-H. Convergence of Block Lanczos Method for Eigenvalue Clusters; 2013.}
\begin{frame}{Convergence Properties}
	\begin{itemize}[<+->]
		\item{The convergence of these subspace algorithms depends on:}
		\begin{itemize}
			\item{\alert{Number of eigenvalues} requested}
			\item{The \alert{block size} of the algorithm}
			\item{The \alert{density} of the eigenvalue spectrum}
			\item{The \alert{initial guess} eigenvectors}
			%\item{The \alert{condition number} of the eigenvector matrix}
			\item{The \alert{preconditioner} (Davidson only)}
		\end{itemize}
		\item{They can \alert{converge to an eigenvalue other than the target} if:}
		\begin{itemize}
			\item{The block size is too small compared to degeneracy}
			\item{The approximate eigenvectors become non-orthogonal}
		\end{itemize}
		\item{My recommendation for guess eigenvectors is
		\begin{eqnarray}
			v_j^{(i)} = normalize\left(\frac{1}{|A_{ii} - A_{jj}| + 1} \right).
		\end{eqnarray}
		}
	\end{itemize}
\end{frame}

\begin{comment}
{%
\setbeamertemplate{frame footer}{Hern, V.; Tom, A.; Vidal, V. SLEPc Technical Reports, 2007.}
\begin{frame}{Orthogonalization}
	\begin{itemize}
		\item{The condition number, $\kappa$, is bound from below by
			\begin{eqnarray}
				\kappa \geq \frac{Max(A_{ii})}{Min(A_{jj})}
			\end{eqnarray}
		}
		\item{The Gram-Schmidt procedure has numerical issues,
			\begin{eqnarray}
				||\mathbf{I} - \mathbf{Q^TQ}|| \leq \frac{\alpha\kappa^2}{1-\beta\kappa^2}.
			\end{eqnarray}
		}
		\item{Modified Gram-Schmidt is better, but not perfect,
			\begin{eqnarray}
				||\mathbf{I} - \mathbf{Q^TQ}|| \leq \frac{\gamma\kappa}{1-\eta\kappa}
			\end{eqnarray}
		}
		\item{May need multiple orthogonalization steps}
	\end{itemize}

\end{frame}
\end{comment}

{%
\setbeamertemplate{frame footer}{Overhauser, A. W. Phys. Rev. Lett. 1960, 4 (9), 462–465.}
\begin{frame}{Project Goal}
	\begin{itemize}[<+->]
		\item{Can we use the HF-Stability analysis to determine where the HEG is unstable?}
		\item{Will this predict the known tendency of crystallization at low density?}
		\item{It is known that the RHF-GHF instability persists at all densities for the HEG$^1$.}
		\item{Can we show this numerically, and discriminate between RHF-UHF and RHF-GHF instability?}
		\item{(Future) Can we combine this with to improve the efficacy of correlation theories (CC, MBPT)?}
	\end{itemize}
\end{frame}

\begin{frame}{First Brillouin Zone}
	\begin{columns}[c] % align columns
		\begin{column}{.4\textwidth}
			\begin{itemize}
				\item {Excite Ony in the X direction}
				\item {-X virtuals matter due to $\vec{k}' = \vec{k} + \vec{G}$}
			\end{itemize}
		\end{column}
		\hfill
		\begin{column}{0.6\textwidth}
		    \begin{overprint}
			    \onslide<1>\includegraphics[width=\linewidth]{../../images/1stBZ.eps}

\begin{frame}{Orbital Energies}
	\begin{columns}[c] % align columns
		\begin{column}{.4\textwidth}
			\begin{itemize}
				\item {Nk = 57 reproduces the orbital energies reasonably well}
				\item {Worst towards $\Gamma$, better for higher $|\vec{k}|$}
			\end{itemize}
		\end{column}
		\hfill
		\begin{column}{0.6\textwidth}
		    \begin{overprint}
			    \onslide<1>\includegraphics[width=\linewidth]{../../images/energycompare.eps}

\begin{frame}{Matrix Size Scaling}
	\begin{columns}[c] % align columns
		\begin{column}{.4\textwidth}
			\begin{itemize}
				\item {Scales as $2N_{exc}$}
				\item {$N_{occ}$ and $N_{vir}$ both scale with $N_{kpoints}^D$}
				\item {Use an iterative subspace method }
				\item {Jacobi-Davidson(-Liu)}
			\end{itemize}
		\end{column}
		\hfill
		\begin{column}{0.6\textwidth}
		    \begin{overprint}
			    \onslide<1>\includegraphics[width=\linewidth]{../../images/matscale.eps}

\begin{frame}{Matrix-Vector Scaling}
	\begin{columns}[c] % align columns
		\begin{column}{.4\textwidth}
			\begin{itemize}
				\item {Scales as $N_{exc} \times  N_{occ}$ due to momentum conservation}
			\end{itemize}
		\end{column}
		\hfill
		\begin{column}{0.6\textwidth}
		    \begin{overprint}
			    \onslide<1>\includegraphics[width=\linewidth]{../../images/mvscaling.eps}

@@

\begin{comment}
\begin{frame}{Algorithm Wall time}
	\begin{columns}[c] % align columns
		\begin{column}{.4\textwidth}
			\begin{itemize}
				\item {Roughly quadratic wrt matrix dimension.}
				\item {Outliers correspond to high condition number}
			\end{itemize}
		\end{column}
		\hfill
		\begin{column}{0.6\textwidth}
		    \includegraphics[width=\linewidth]{../../images/overallruntime.eps}
		\end{column}
	\end{columns}
\end{frame}
\end{comment}

\begin{frame}{Efficacy of Davidson's Algorithm}
	\begin{columns}[c] % align columns
		\begin{column}{.4\textwidth}
			\begin{itemize}
				\item {Reproduces Exact result to machine precision in all test cases.}
				\item {Odd spikes are due to approximating circle by squares}
			\end{itemize}
		\end{column}
		\hfill
		\begin{column}{0.6\textwidth}
		    \begin{overprint}
			    \onslide<1>\includegraphics[width=\linewidth]{../../images/dav_vs_exact.eps}

@@

{%
\setbeamertemplate{frame footer}{Bernu, B.; Delyon, F.; Holzmann, M.; Baguet, L. Phys. Rev. B 2011, 84 (11), 115115.}
\begin{frame}{Dependence on r${}_s$}
	Crossover from stable to unstable in 2D agrees with previous results.
	\includegraphics[width=.95\linewidth]{../../images/MeVsBernu.pdf}
\end{frame}

{%
\setbeamertemplate{frame footer}{}
\begin{frame}{Dependence on r${}_s$}
\centering
	3D data implies some stability at high density 
	\includegraphics[width=.65\linewidth]{../../images/3dstab.pdf}
\end{frame}

{%
\setbeamertemplate{frame footer}{}
\begin{frame}{Dependence on r${}_s$}
  \centering
	Preliminary data for 1D shows some stability 
	\includegraphics[width=.65\linewidth]{../../images/1dstab.pdf}
\end{frame}

\begin{comment}
\section{Concluding Remarks}
{%
\setbeamertemplate{frame footer}{}
\begin{frame}{Next Steps}
	\resizebox{\textwidth}{!}{%
	\begin{tabular}{l c c}
		Objective                 & Human Time   & Computer Time    \\
		\midrule
		Other instabilities in 2D & 30 mins & Week \\
		All 3D & $\frac{1}{2}$ Day&  $>$Weeks? Parallel?
	\end{tabular}}
\end{frame}
}
\end{comment}

%{%
%\setbeamertemplate{frame footer}{}
%\begin{frame}[fragile]{Backup slide}
%	Empty Backup
%\end{frame}
%
\end{document}

>>>>

\title{Hartree-Fock Stability of the Electron Gas}
% \subtitle{Applied to the Electron Gas}
\date{\today}
\author{Evan Curtin}
\institute{University of Illinois at Urbana-Champaign}
\titlegraphic{\hfill\includegraphics[height=1.5cm]{../../images/imark_bold.eps}}

@@

  \begin{frame}{Matrix Factorizations}
  \vspace{-6mm}
  \begin{eqnarray*}
    \begin{bmatrix}
      \mathbf{A} & \mathbf{B} \\
      \mathbf{B^*} & \mathbf{A^*} 
    \end{bmatrix}
    \begin{bmatrix}
      \mathbf{d} \\
      \mathbf{d^*} 
    \end{bmatrix}
    = 2E_2
    \begin{bmatrix}
      \mathbf{d} \\
      \mathbf{d^*} 
    \end{bmatrix}
  \end{eqnarray*}
  We can now apply the similarity transform defined by the Unitary matrix 
  \begin{eqnarray*}
    \mathbf{U} = \frac{1}{\sqrt{2}}
    \begin{bmatrix}
      \mathbf{I} & -\mathbf{I} \\
      \mathbf{I} & \mathbf{I} 
    \end{bmatrix}
  \end{eqnarray*}
  after which the transformed eigenvalue problem has the form
  \begin{eqnarray*}
    \frac{1}{2}
    \begin{bmatrix}
      \mathbf{A + B + A^* + B^*} & \mathbf{-A + A^* + B - B^*} \\
      \mathbf{-A + A^* - B + B^*} & \mathbf{A^* + A - B - B^*} 
    \end{bmatrix}
    \begin{bmatrix}
      \mathbf{d + d^*} \\
      \mathbf{d - d^*} 
    \end{bmatrix}
    &=& 2E_2
    \begin{bmatrix}
      \mathbf{d + d^*} \\
      \mathbf{-d + d^*} 
    \end{bmatrix} \\
    &=& 2E_2
    \begin{bmatrix}
      \mathbf{Re(d)} \\
      \mathbf{Im(d)} 
    \end{bmatrix}
  \end{eqnarray*}
  If $\mathbf{A}$ and $\mathbf{B}$ are both real, $\mathbf{A = A^*}$ and $\mathbf{B = B^*}$ and the 
  above simplifies to 
  \begin{eqnarray*}
    \begin{bmatrix}
      \mathbf{A + B} & \mathbf{0} \\
      \mathbf{0} & \mathbf{A - B} 
    \end{bmatrix}
    \begin{bmatrix}
      \mathbf{Re(d)} \\
      \mathbf{Im(d)} 
    \end{bmatrix}
    &=& 2E_2   
    \begin{bmatrix}
      \mathbf{Re(d)} \\
      \mathbf{Im(d)} 
    \end{bmatrix}
  \end{eqnarray*}
  \end{frame}
\end{comment}

{%
	\setbeamertemplate{frame footer}{1) Seeger, R.; Pople, J. A. J. Chem. Phys. 1977, 66 (7), 3045.}
\begin{frame}{Hartree-Fock Stability Conditions}
	\begin{itemize}
		\item[]{The Matrix Equation Factorizes} \\~\\
		\item[]{
			\resizebox{0.9\textwidth}{!}{%
			\begin{tabular}{r@{\qquad}cccccc}
				  \multirow{1}{*}{Solution Type} & \multicolumn{6}{c}{Space Type}  \\
				  \cmidrule{2-7}
				  & Real RHF & Complex RHF & Real UHF & Complex UHF & Real GHF & Complex GHF \\
				  \midrule
				  Real RHF    & ${}^1\mathbf{A}' + {}^1\mathbf{B}'$ & ${}^1\mathbf{A}' - {}^1\mathbf{B}'$ & 
				  ${}^3\mathbf{A}' + {}^3\mathbf{B}'$ & ${}^3\mathbf{A}' - {}^3\mathbf{B}'$ & 
				  ${}^3\mathbf{A}' + {}^3\mathbf{B}'$ & ${}^3\mathbf{A}' - {}^3\mathbf{B}'$  \\
				  Complex RHF & - & \alert{${}^1\mathbf{H}'$} & - & \alert{${}^3\mathbf{H}'$} & - & 
				  ${}^3\mathbf{H}'$ \\
				  Real UHF    & - & - & $\mathbf{A}' + \mathbf{B}'$ & $\mathbf{A}' - \mathbf{B}'$ & 
				  $\mathbf{A}'' + \mathbf{B}''$ & $\mathbf{A}'' - \mathbf{B}''$ \\
				  Complex UHF & - & - & - &  $\mathbf{H}'$ & - & $\mathbf{H}'$ \\
				  Real GHF    & - & - & - & - & $\mathbf{A} - \mathbf{B}$ &  $\mathbf{A} - \mathbf{B}$  \\
				  Complex GHF & - & - & - & - & - & $\mathbf{H}$ \\
				  \bottomrule
			\end{tabular}}
		\item[] \tiny Table reproduced from Seeger \& Pople$^1$
		}
	\end{itemize}
\end{frame}

@@

{%
\setbeamertemplate{frame footer}{Overhauser, A. W. Phys. Rev. Lett. 1960, 4 (9), 462–465.}
\begin{frame}{Project Goal}
	\begin{itemize}[<+->]
		\item{Can we use the HF-Stability analysis to determine where the HEG is unstable?}
		\item{Will this predict the known tendency of crystallization at low density?}
		\item{It is known that the GHF solution persists at all densities for the HEG$^1$.}
		\item{Can we show this numerically?}
		\item{(Future) Can we combine this with to improve the efficacy of correlation theories (CC, MBPT)?}
	\end{itemize}
\end{frame}

\begin{frame}{First Brillouin Zone}
	\begin{columns}[c] % align columns
		\begin{column}{.4\textwidth}
			\begin{itemize}
				\item {Excite Ony in one direction}
				\item {Consider only First BZ due to $\vec{k}' = \vec{k} + \vec{G}$}
			\end{itemize}
		\end{column}
		\hfill
		\begin{column}{0.6\textwidth}
		    \begin{overprint}
			    \onslide<1>\includegraphics[width=\linewidth]{../../images/1stBZ.eps}

\begin{frame}{Orbital Energies}
	\begin{columns}[c] % align columns
		\begin{column}{.4\textwidth}
			\begin{itemize}
				\item {57 grid points per dimension reproduces the orbital energies reasonably well}
				\item {Worst towards $\Gamma$, better for higher $|\vec{k}|$}
			\end{itemize}
		\end{column}
		\hfill
		\begin{column}{0.6\textwidth}
		    \begin{overprint}
			    \onslide<1>\includegraphics[width=\linewidth]{../../images/energycompare.eps}

			\end{overprint}
		\end{column}
	\end{columns}
\end{frame}
\end{comment}

\begin{frame}{Convergence of Stability Curves}
	\includegraphics[width=.95\linewidth]{../../images/singlet_onset.eps}
	\\
	\includegraphics[width=.95\linewidth]{../../images/triplet_onset.eps}
\end{frame}

{%
\setbeamertemplate{frame footer}{Bernu, B.; Delyon, F.; Holzmann, M.; Baguet, L. Phys. Rev. B 2011, 84 (11), 115115.}
\begin{frame}{Dependence on r${}_s$}
	Previously reported transition in 2d: $rs \approx 0.8 $
  \begin{center}
  	\includegraphics[width=.6\linewidth]{../../images/Bernu_2dEnergyGain.pdf}
	\end{center}
\end{frame}

\begin{frame}{Final Stability Curves}
  \vspace{-0.5in}
	\includegraphics[width=\linewidth]{../../images/stability.pdf}
\end{frame}

\section{Concluding Remarks}
{%
\setbeamertemplate{frame footer}{}
\begin{frame}{Conclusions}
  \begin{itemize}
    \item {Numerical studies likely support Overhauser's theorem in 1D}
    \item {Stability corresponds with direct phase calculations in 2D}
    \item {In 3D, predicts transition at higher density than is currently known via
    other methods.}
  \end{itemize}
\end{frame}
}

{%
\setbeamertemplate{frame footer}{}
\begin{frame}{Next Steps}
  \begin{itemize}
    \item {Directly calculate more exotic spin phases of HEG within HF theory}
    \item {Use those as starting points for CC, MBPT theories}
    \item {Apply a similar approach to finite molecular systems}
  \end{itemize}
\end{frame}
}

@@

 2. Li, R.-C.; Zhang, L.-H. Convergence of Block Lanczos Method for Eigenvalue Clusters; 2013.}
\begin{frame}{Convergence Properties}
	\begin{itemize}[<+->]
		\item{The convergence of these subspace algorithms depends on:}
		\begin{itemize}
			\item{\alert{Number of eigenvalues} requested}
			\item{The \alert{block size} of the algorithm}
			\item{The \alert{density} of the eigenvalue spectrum}
			\item{The \alert{initial guess} eigenvectors}
			\item{The \alert{preconditioner} (Davidson only)}
		\end{itemize}
		\item{They can \alert{have convergence issues} if:}
		\begin{itemize}
			\item{The block size is too small compared to degeneracy}
			\item{The approximate eigenvectors become non-orthogonal}
		\end{itemize}
		\item{My recommendation for guess eigenvectors is
		\begin{eqnarray}
			v_j^{(i)} = normalize\left(\frac{1}{|A_{ii} - A_{jj}| + 1} \right).
		\end{eqnarray}
		}
	\end{itemize}
\end{frame}

@@

{%
\setbeamertemplate{frame footer}{}
\begin{frame}{Efficacy of Davidson's Algorithm}
	\begin{columns}[c] % align columns
		\begin{column}{.4\textwidth}
			\begin{itemize}
				\item {Reproduces Exact result to machine precision in all test cases.}
				\item {Odd spikes are due to approximating circle by squares}
			\end{itemize}
		\end{column}
		\hfill
		\begin{column}{0.6\textwidth}
		    \begin{overprint}
			    \onslide<1>\includegraphics[width=\linewidth]{../../images/dav_vs_exact.eps}

@@

{%
\setbeamertemplate{frame footer}{Hern, V.; Tom, A.; Vidal, V. SLEPc Technical Reports, 2007.}
\begin{frame}{Orthogonalization}
	\begin{itemize}
		\item{The condition number, $\kappa$, is bound from below by
			\begin{eqnarray}
				\kappa \geq \frac{Max(A_{ii})}{Min(A_{jj})}
			\end{eqnarray}
		}
		\item{The Gram-Schmidt procedure has numerical issues,
			\begin{eqnarray}
				||\mathbf{I} - \mathbf{Q^TQ}|| \leq \frac{\alpha\kappa^2}{1-\beta\kappa^2}.
			\end{eqnarray}
		}
		\item{Modified Gram-Schmidt is better, but not perfect,
			\begin{eqnarray}
				||\mathbf{I} - \mathbf{Q^TQ}|| \leq \frac{\gamma\kappa}{1-\eta\kappa}
			\end{eqnarray}
		}
		\item{May need multiple orthogonalization steps}
	\end{itemize}

\end{frame}

\end{document}
